MSDS 7333 Case Study 06

### 2017-06-19 MSDS7333_CaseStudy06_MBaldree_BBrock_TElkins_AKelly.Rmd
Added meat to the bones

### Assignment: 
1. Conduct a more thorough data analysis into these two MAC addresses, including determining locations by using data corresponding to both: 
    + chosenAP = "00:0f:a3:39:e1:c0"
    + rejectedAP = "00:0f:a3:39:dd:cd"
* *Which of these two MAC addresses should be used and which should not be used for Real Time Location System (RTLS)?*
* *Which MAC address yields the best prediction of location?including determining locations by using data corresponding to both MAC addresses*
* *Does using data for both MAC addresses simultaneously yield more, or less accurate prediction of location?*

@@

2. Implement alternative k-nearest precition method using weights on received signal strength.
* *For what range of values of weights are you able to obtain better prediction values than for the unweighted approach?*
* *Use calcError() to compare approaches*
<font color = "green">

## Introduction
#### *Explain the case study*
- Discuss the goal of the case study

@@

- Discuss the different Access points and MAC Addresses

@@

  - Discuss why we're trying to remove one

@@

- BRIEF overview of K-Nearest Neighbors

@@

- Discuss the alternative weighted method.

@@

- Discuss potential for prediction and what we expect to come of this.

@@

</font>

@@

## The Approach (Explained)

### Part 1

Before we begin, we will need to set the parameters for printing and formatting
```{r}
options(digits = 2)
options(error = recover, warn = 2)
```
Before continuing, we will need to address another file which houses the predefined functions. This helps to save notebook real estate and prevents confusion:
```{r}
source("MSDS7333-baldree-case6-fx.r", print.eval = TRUE)
```

@@

* *Which of these two MAC addresses should be used and which should not be used for Real Time Location System (RTLS)?*

@@

```{r, fig.height=10}
df = subset(offline, posX == 24 & posY == 4 & !(mac %in% remainingAP), c(signal, angle, mac))
plotDensitySignalStrength(df)
```

From this plot, it can be seen that the rejectedAP has a higher signal density at lower strength while the acceptedAP has a higher density coverage at a higher strength. Judging by the distributions themselves, it seems as if many of the plots are normally distributed while some of the other plots (i.e. "dd:cd" at 180 and "e1:c0" at 315) are heavily skewed or even bimodal. Looking at each AP from top to bottom, there seems to exist an obvious pattern with respect to angle. This would suggest the median of each distribution varies by angle and attributes to a difference in perceived strength.

This evidence would suggest both addresses are kept within the analysis to increase the coverage for the entire spectrum of strengths.

@@

### <font color = "red"> &darr; this section may no longer be needed. Now missing from the original code. &darr;</font>  
```{r}
#offline$posXY = paste(offline$posX, offline$posY, sep="-")
#
# create data frames for each combination
#byLocAngleAP = with(offline, by(offline, list(posXY, angle, mac), function(x) x))
```

Now, we calculate the summary statistics:
```{r}
#signalSummary = lapply(byLocAngleAP, function(oneLoc) {
#  ans = oneLoc[1, ]
#  ans$medSignal = median(oneLoc$signal)
#  ans$avgSignal = mean(oneLoc$signal)
#  ans$num = length(oneLoc$signal)
#  ans$sdSignal = sd(oneLoc$signal)
#  ans$iqrSignal = IQR(oneLoc$signal)
#  ans
#})
#offlineSummary = do.call("rbind", signalSummary)
```
### <font color = "red"> &uarr; this section may no longer be needed. Now missing from the original code. &uarr;</font>

@@

@@

```{r}
source("MSDS7333-baldree-case6-fx.r", print.eval = TRUE)
```

@@

*** 
#Part 2: Alternatvie k-nearest neighbors
```{r}
# create weights for neighbors based on signal strength and multiple times X and Y and sum neighbors
# to get X and Y.
estXY = predXY(newSignals = onlineSummary[ , 6:12],
               newAngles = onlineSummary[ , 4],
               offlineSummary, numAngles = 3, k = 5, weighted=TRUE)
actXY = onlineSummary[ , c("posX", "posY")]
calcError(estXY, actXY)
```

>>>>

### Table of Contents:
<li>[Assignment Requirements](#AssignmentOutline)</li>
<li>[Introduction](#Introduction)</li>
<li>[Part 1](#Part1)</li>
  <li>[Signal Collection](#SigCol)</li>
  <li>[Signal Strength](#SigStren)</li>
  <li>[Signal Density](#SigDen)</li>
  <li>[Discussion](#Discussion)</li>
<li>[Part2](#Part 2)</li>

<li>[References](#References)</li>
<li>[Appendix](#Appendix)</li>

<a id='AssignmentOutline'>

### <b>Assignment</b> </a>
1. Conduct a more thorough data analysis into these two MAC addresses, including determining locations by using data corresponding to both the chosen and rejected Access Points:
    + *Which of these two MAC addresses should be used and which should not be used for Real Time Location System (RTLS)?*
    + *Which MAC address yields the best prediction of location? (Including determining locations by using data corresponding to both MAC addresses)*
    + *Does using data for both MAC addresses simultaneously yield more, or less accurate prediction of location?*

2. Implement alternative k-nearest precition method using weights on received signal strength.
    + *For what range of values of weights are you able to obtain better prediction values than for the unweighted approach?*
    + *Use calcError() to compare approaches*

@@

<a id='Introduction'>

### <b>Introduction</b></a>  
#### *Explain the case study*

<font color = "purple"><b>>>>>Everything in this color to be removed at final review<<<<</b></font>

<font color = "purple"><b>
- Discuss the goal of the case study
</b></font>

@@

<font color = "purple"><b>
- Discuss the different Access points and MAC Addresses
</b></font>

@@

<font color = "purple"><b>
- Discuss why we're trying to remove one
</b></font>

@@

<font color = "purple"><b>
- BRIEF overview of K-Nearest Neighbors
</b></font>

@@

<font color = "purple"><b>
- Discuss the alternative weighted method. 
</b></font>

@@

<font color = "purple"><b>
- Discuss potential for prediction and what we expect to come of this.
</b></font>

@@

@@

### Part 1
#### The Approach (Explained)

Before we begin, we will need to set the parameters for printing and formatting
```{r}
options(digits = 2)
options(error = recover, warn = 2)
```

Before continuing, we will need to address another file which houses the predefined functions. This helps to save notebook real estate and prevents confusion:
```{r}
source("MSDS7333-baldree-case6-fx.r", print.eval = TRUE)
```

@@

#### * *Which of these two MAC addresses should be used and which should not be used for Real Time Location System (RTLS)?*

@@

```{r, fig.height=10}
df = subset(offline, posX == 24 & posY == 4 & !(mac %in% remainingAP), c(signal, angle, mac))
plotDensitySignalStrength(df)
```
From this plot, it can be seen that the rejectedAP has a higher signal density at lower strength while the acceptedAP has a higher density coverage at a higher strength. Judging by the distributions themselves, it seems as if many of the plots are normally distributed while some of the other plots (i.e. "dd:cd" at 180 and "e1:c0" at 315) are heavily skewed or even bimodal. Looking at each AP from top to bottom, there seems to exist an obvious pattern with respect to angle. This would suggest the median of each distribution varies by angle and attributes to a difference in perceived strength.

@@

```{r}
offline$posXY = paste(offline$posX, offline$posY, sep="-")

# create data frames for each combination
byLocAngleAP = with(offline, by(offline, list(posXY, angle, mac), function(x) x))
```

Now, we calculate the summary statistics:
```{r}
signalSummary = lapply(byLocAngleAP, function(oneLoc) {
  ans = oneLoc[1, ]
  ans$medSignal = median(oneLoc$signal)
  ans$avgSignal = mean(oneLoc$signal)
  ans$num = length(oneLoc$signal)
  ans$sdSignal = sd(oneLoc$signal)
  ans$iqrSignal = IQR(oneLoc$signal)
  ans
})
offlineSummary = do.call("rbind", signalSummary)
```

@@

#### We now have enough information to address these questions before an analysis of kNN is conducted:

* **Which MAC address yields the best prediction of location? (Include determining locations by using data corresponding to both MAC addresses)**
* **Does using data for both MAC addresses simultaneously yield a more or less accurate prediction of location?**

The evidence would suggest both addresses should be kept within the analysis to increase the coverage for the entire spectrum of strengths. The address chosen to be rejected not only has a lower signal strength when measured from either end of the building, it also shows to have a higher signal density when it is further away from the receiver. The address chosen to be kept has a higher signal density closer to the receivers. By keeping both addresses, the model should be able to generate a more accurate prediction of the layout of the building. This theory will be tested as we tour through k-Nearest Neighbors.

@@

@@

*** 
<a id='Part2'>

#Part 2: Alternatvie k-Nearest Neighbors
</a>
```{r}
# create weights for neighbors based on signal strength and multiple times X and Y and sum neighbors
# to get X and Y.
estXY = predXY(newSignals = onlineSummary[ , 6:12],
               newAngles = onlineSummary[ , 4],
               offlineSummary, numAngles = 3, k = 5, weighted=TRUE)
actXY = onlineSummary[ , c("posX", "posY")]
calcError(estXY, actXY)
```

***
<a id='References'>

### References  
</a>

<a id='Appendix'>

### Appendix (including function code)  
</a>
This is the code used to define the functions used within the study. It can be found under 'MSDS7333-baldree-case6-fx.r' which is included in the final submission  
```{r}
processLine = function(x)
{
  # Process a line of text returning a 11x10 matrix of tokens
  #
  # Args:
  #   x: The line of text
  #
  tokens = strsplit(x, "[;=,]")[[1]]

  # return NULL if only 10 tokens
  if (length(tokens) == 10)
    return(NULL)

  tmp = matrix(tokens[-(1:10)], ncol = 4, byrow = TRUE)
  mat = cbind(matrix(tokens[c(2, 4, 6:8, 10)], nrow(tmp), 6, byrow = TRUE), tmp)
}

roundOrientation = function(angles) {
  # Round an array of degree angles to the nearest 45 degree angle.
  #
  # Args:
  #   angles: The line of text
  #
  refs = seq(0, by = 45, length  = 9)
  q = sapply(angles, function(o) which.min(abs(o - refs)))
  c(refs[1:8], 0)[q]
}

readData = function(filename = 'offline.final.trace.txt',
                    subMacs = c("00:0f:a3:39:e1:c0", "00:0f:a3:39:dd:cd", "00:14:bf:b1:97:8a",
                                "00:14:bf:3b:c7:c6", "00:14:bf:b1:97:90", "00:14:bf:b1:97:8d",
                                "00:14:bf:b1:97:81")) {
  # Process the csv file peforming the following steps:
  #   Read all non-comment lines.
  #   Call processLine() for each line of text.
  #   Convert list of text into a data frame
  #   Name columns of data frame
  #   Define column types
  #   Convert millisecond time to seconds
  #   Round angle orientations
  #   Drop data
  #
  # Args:
  #   filename: The name of the csv file to process. The default is 'offline.final.trace.txt.'
  #   subMacs: The list of access point MAC addresses to filter the dataset down to. A default list is provided.
  #
  # Returns:
  #   The data frame
  #

  # process csv into matrix of tokens
  txt = readLines(filename)
  lines = txt[substr(txt, 1, 1) != "#"]
  tmp = lapply(lines, processLine)
  # convert into tokens into a data frame
  df = as.data.frame(do.call("rbind", tmp), stringsAsFactors = FALSE)

  # variable names
  names(df) = c("time", "scanMac", "posX", "posY", "posZ", "orientation", "mac", "signal", "channel", "type")

  # convert numeric values
  numVars = c("time", "posX", "posY", "orientation", "signal")
  df[numVars] = lapply(df[numVars], as.numeric)

  # convert time to POSIX
  df$rawTime = df$time
  df$time = df$time / 1000
  class(df$time) = c("POSIXt", "POSIXct")

  # round orientations to nearest 45
  df$angle = roundOrientation(df$orientation)

  # keep only signals from access points
  df = df[df$type == "3",]
  df = df[, "type" != names(df)]

  # drop scanMac, posZ, channel, and type - no info in them
  df = df[, !(names(df) %in% c("scanMac", "posZ"))]

  # drop more unwanted access points
  df = df[df$mac %in% subMacs,]

  # drop channel
  df = df[, "channel" != names(df)]

  return(df)
}

plotBoxplotSignalStrength = function(df) {
  # Box plot signal strength
  #
  # Args:
  #   df: data frame of data to analyze
  #
  oldPar = par(mar = c(3.1, 3.1, 1, 1), mfrow = c(1,1))
  par(mai=c(.8, .8, .5, .25))
  library (lattice)

  print(bwplot(signal ~ factor(angle) | mac, data = df, layout = c(2,1),
               main="Signal Strength Distribution for Access Points",
               xlab="Angle Measured (deg)", ylab="Signal Strengh (dBM)"))
  par(oldPar)
}

plotDensitySignalStrength = function(df) {
  # Density plot signal strength
  #
  # Args:
  #   df: data frame of data to analyze
  #
  oldPar = par(mar = c(3.1, 3.1, 1, 1), mfrow = c(1,1))
  par(mai=c(.8, .8, .5, .25))
  library (lattice)

  print(densityplot( ~ signal | mac + factor(angle), data = df, bw = 0.5, plot.points = FALSE,
        main="Signal Strength Distribution for Access Points",
        xlab="Angle Measured (deg)", ylab="Signal Strengh (dBM)"))
  par(oldPar)
}

plotStdDevSignalStrength = function(df) {
  # Standard Deviation of signal strength average
  #
  # Args:
  #   df: data frame of data to analyze
  #
  oldPar = par(mar = c(3.1, 3.1, 1, 1), mfrow = c(1,1))
  par(mai=c(.8, .8, .5, .25))
  library (lattice)
  breaks = seq(-90, -30, by=5)

  print(bwplot(sdSignal ~ cut(avgSignal, breaks=breaks) | mac, data=df,
               xlab = "Mean Signal (dBM)", ylab = "Signal Standard Deviation (dBM)",
               main = "Standard Deviation of Average Signal Strength from Detector to Access Point",
               cex.main=.8, cex.axis=.8, cex.lab=.8, layout = c(1, 2)))
  par(oldPar)
}

plotSignalMaps = function(macAddresses) {
  # Plot the number of signals per XY location at XY location.
  #
  # Args:
  #   macAddresses: a list of macs to plot signals
  #
  oldPar = par(mar = c(3.1, 3.1, 1, 1), mfrow = c(1,1))
  par(mfrow = c(2, 1), mai=c(.8, .8, .4, .25))

  for (i in macAddresses) {
    # filter data set
    df = offline[offline$mac %in% i, ]
    # location of datapoints filtered
    locDF = with(df, by(df, list(posX, posY), function(x) x))
    # drop locations that were not observed, null
    locDF = locDF[!sapply(locDF, is.null)]

    # determine the number of observations recorded at each location
    locCounts = sapply(locDF, nrow)

    # keep position information with location
    locCounts = sapply(locDF, function(df) c(df[1, c("posX", "posY")], count = nrow(df)))

    # transpose matrix
    locCounts = t(locCounts)
    plot(locCounts, type = "n", xlab = "Position X", ylab = "Position Y",
         main = paste("Total Signals from Detector to Access Point: ", i),
         cex.main=.8, cex.axis=.8, cex.lab=.8)
    text(locCounts, labels = locCounts[, 3], cex = .6, srt = 45)
  }
  par(oldPar)
}

castOnline = function(df) {
  # Cast dataframe of online data so each access point (AP) is in its own column.
  # Matrix is now a 1x7 since we are including rejected AP.
  #
  # Args:
  #   df: online data frame
  #
  keepVars = c("posXY", "posX","posY", "orientation", "angle")
  numOfAPs = length(unique(df$mac))
  byLoc = with(df,
               by(df, list(posXY),
                  function(x) {
                    ans = x[1, keepVars]
                    avgSS = tapply(x$signal, x$mac, mean)
                    y = matrix(avgSS, nrow = 1, ncol = numOfAPs,
                               dimnames = list(ans$posXY, names(avgSS)))
                    cbind(ans, y)
                  }))
  return(do.call("rbind", byLoc))
}

createOfflineSummary = function(df) {
  # In order to examine distribution for all locations, angles, and are two interested APs, we
  # will create a summary statistics for all location-orientation-AP combinations with a new factor.
  # For each combination there are around 100 observations.
  #
  # Args:
  #   df: offline data frame
  #
  df$posXY = paste(df$posX, df$posY, sep="-")

  # create data frames for each combination
  byLocAngleAP = with(df, by(df, list(posXY, angle, mac), function (x) x))

  # summary statistic
  signalSummary = lapply(byLocAngleAP, function(oneLoc) {
    ans = oneLoc[1, ]
    ans$medSignal = median(oneLoc$signal)
    ans$avgSignal = mean(oneLoc$signal)
    ans$num = length(oneLoc$signal)
    ans$sdSignal = sd(oneLoc$signal)
    ans$iqrSignal = IQR(oneLoc$signal)
    ans
  })
  return(do.call("rbind", signalSummary))
}

reshapeSS = function(df,
                     varSignal = "signal",
                     keepVars = c("posXY", "posX", "posY"),
                     sampleAngle = FALSE,
                     refs = seq(0, 315, by=45)) {
  # reshape signal strength help function
  # aggregate signal strengths from these angles and create a data structure similar to onlineSummary
  # Args:
  #   df: dataframe to reshape
  #   varSignal: variable to aggregate
  #   keepVars: variables to retain
  #   sampleAngle: true if we want to select one angle at random from df
  #   refs: angle references
  #
  numOfAPs = length(unique(df$mac))
  byLocation = with(df, by(df, list(posXY),
                             function(x) {
                               # select one angle at random for each location
                               if (sampleAngle) {
                                 x = x[x$angle == sample(refs, size = 1), ]}
                               ans = x[1, keepVars]
                               avgSS = tapply(x[, varSignal], x$mac, mean)
                               y = matrix(avgSS, nrow = 1, ncol = numOfAPs,
                                          dimnames = list(ans$posXY, names(avgSS)))
                               cbind(ans, y)
                             }))
  return(do.call("rbind", byLocation))
}

selectTraininingData = function(newObservationAngle, df = NULL, m = 1){
  # select observations from df to analyze aggregate signal strengths
  # from these angles and create a data structure similar to onlineSummary.
  #
  # Args:
  #   newObservationAngle: the angle of the new observation
  #   df: offline summary data frame
  #   m: number, between 1 and 5, of angles to keep

  # m is the number of angles to keep between 1 and 5
  refs = seq(0, by = 45, length  = 8)
  nearestAngle = roundOrientation(newObservationAngle)

  # handle odd and even m number
  if (m %% 2 == 1)
    angles = seq(-45 * (m - 1) /2, 45 * (m - 1) /2, length = m)
  else {
    m = m + 1
    angles = seq(-45 * (m - 1) /2, 45 * (m - 1) /2, length = m)
    if (sign(newObservationAngle - nearestAngle) > -1)
      angles = angles[ -1 ]
    else
      angles = angles[ -m ]
  }
  # map angles to values in refs
  # negative angles and angles greater than 360 are mapped to appropriate angles; e.g., -45 maps to 335 and 405 maps to 45
  angles = angles + nearestAngle
  angles[angles < 0] = angles[ angles < 0 ] + 360
  angles[angles > 360] = angles[ angles > 360 ] - 360
  angles = sort(angles)
  # select observations to analyze
  subset = df[df$angle %in% angles, ]
  # reshape signal strength
  return(reshapeSS(subset, varSignal = "avgSignal"))
}

findNN = function(newSignal, trainingSubset) {
  # want to look the distance in terms of signal strengths from these training data to the new
  # data point. we need to calculate teh distrance from the new point to all observations in the
  # training set with findNN().
  # returns locations of the training observations in order of closeness to the new observation's signal strength.
  #
  # Args:
  #   newSignal: signal of new observation
  #   trainingSubset: training data to find neighbors
  # Returns:
  #   training neighbors
  #
  cols = length(trainingSubset)
  diffs = apply(trainingSubset[ , 4:cols], 1, function(x) x - newSignal)
  dists = apply(diffs, 2, function(x) sqrt(sum(x^2)) )
  #closest = order(dists)
  closest = cbind(trainingSubset[, 1:3 ], dists)
  return(closest[order(dists),])
  #return(trainingSubset[closest, 1:3 ])
}

predXY = function(newSignals, newAngles, training, numAngles = 1, k = 3, weighted=FALSE){
  # Predict the XY coordinates given a list of signals along with their angles measured and training data.
  # k neighbors will be found to make the prediction of coordinate.
  #
  # Args:
  #   newSignals: list of signals to predict
  #   newAngles: list of angles for signals to predict
  #   trainingData: data needed for k-nn
  #   numAngles: angles we want to use for prediction
  #   k: number of neighbors to use for prediction
  #   weighted: if TRUE, distances to neighbors will be used to weight XY for each neighbor to determine
  #             location.
  #
  closeXY = list(length = nrow(newSignals))

  for (i in 1:nrow(newSignals)) {
    trainSS = selectTraininingData(newAngles[i], training, m = numAngles)
    closeXY[[i]] = findNN(newSignal = as.numeric(newSignals[i, ]), trainSS)
  }
  # for some k of nearest neighbors, simply average the first k locations
  # could have used weights in the average that are inversely proportional to the distance in signal strength
  # from the test observation. will need to call findNN() to get distance to point. the weights
  # are 1/d / sum(1/d).
  # could also use a different metric besides Euclidean like Manhattan.
  # could use medians instead of averages when combining neighbors if the distribution of values are quite skewed.
  if (!weighted) {
    estXY = lapply(closeXY, function(x) sapply(x[ , 2:3], function(x) mean(x[1:k])))
  } else {
    # sum X and Y weighted neighbor values.
    estXY = lapply(closeXY, function(x) {
      wt = (1/x[1:k, 4])/sum(1/x[1:k, 4])
      xy = cbind(x[1:k,], wt)
      # sum weighted neighbors
      return(colSums(xy[1:4, 2:3] * xy[1:4, 4]))
    })
  }
  return(do.call("rbind", estXY))
}

calcError = function(estXY, actualXY) {
  # compare fit numerically with sum of squared errors
  #
  # Args:
  #   estXY:
  #   actualXY:
  # Returns:
  #   calculated residual sum of squares
  return(sum(rowSums((estXY - actualXY) ^ 2)))
}

plotSSErrors = function(err, K){
  # Plot results as sum of squared errors as function of k
  #
  # Args:
  #   err: Error array
  #   K: number of neighbors
  #

  oldPar = par(mar = c(3.1, 3.1, 1, 1), mfrow = c(1,1))
  par(mai=c(.8, .8, .4, .25))

  plot(y = err, x = (1:K),  type = "l", lwd= 2, ylim = c(900, 2100), cex.main=.8, cex.axis=.8, cex.lab=.8,
       xlab = "Number of Neighbors",
       ylab = "Sum of Squared Errors",
       main = "Cross Validation Section of k")

  rmseMin = min(err)
  kMin = which(err == rmseMin)[1]
  segments(x0 = 0, x1 = kMin, y0 = rmseMin, col = gray(0.4), lty = 2, lwd = 2)
  segments(x0 = kMin, x1 = kMin, y0 = 800,  y1 = rmseMin, col = grey(0.4), lty = 2, lwd = 2)
  text(x = kMin - 2, y = rmseMin + 40, label = as.character(round(rmseMin)), col = grey(0.4), cex = .6)

  par(oldPar)
}

wt_signals <- function(ByAvgSig){

  # Name: wt_signals
  # Create weighted data
  #
  # Args:
  #   ByAvgSig:    estimate XY line segment
  #   
  # Returns:
  #   log_ang by weighted_signals matrix

  # calculate signals
  signals <- select(ByAvgSig, -posXY, -posX, -posY, -angle)

  # calculate log_ang
  loc_ang <- select(ByAvgSig, posXY, posX, posY, angle)

  # calculate weighted_signals 
  weighted_signals <- (1/signals) / rowSums(1/signals)

  return(cbind(loc_ang, weighted_signals))
}
```
